Covariate analysis of ROSMAP reprocessed counts (with TMM normalisation and covariates adjusted)

### 2017-03-20 code/Rmd/ROSMAP_REPROCESSED_TMM.Rmd
Added markdown to process ROSMAP reprocessed RNASeq with TMM and SVA normalaisation

---
title: "Covariate analysis of ROSMAP reprocessed counts (with CQN normalisation and covariates adjusted)"
author: "Thanneer Perumal"
date: "`r date()`"
output: html_document
---
```{r knit2synapse, eval=FALSE}
library(synapseClient)
library(knit2synapse) # get the package from devtools::install_github('Sage-Bionetworks/knit2synapse')

@@

knit2synapse::knitToFolderEntity(file = "./ROSMAP_REPROCESSED.Rmd",
                                 parentId = "syn8403743",
                                 entityName = 'ROSMAP Reprocessed RNASeq (Normalised)')
```

@@

library(ggplot2)
library(reshape2)
library(limma)
library(gplots)
library(WGCNA)
library(psych)
library(edgeR)
library(biomaRt)
library(RColorBrewer)
library(cqn)

@@

thisFileName <- 'ROSMAP_REPROCESSED.Rmd'

@@

# Sex
vcd::mosaic(~ msex + cogdx, data = COVARIATES)
```
### CPM Normalisation
Preprocess counts matrix using CQN (Conditional Quantile Normlaisation accounting for both GC content and gene length)

@@

### Normalisation (with NULL)
Initial normalisation usign cqn and dispersion control with voom (with NULL design)
```{r initial.voom.normalisation, fig.height=5, fig.width=5}
# Compute offset for gene length and gc content
CQN.GENE_EXPRESSION = cqn(PROCESSED_COUNTS$filteredExprMatrix$counts, 
                          x = GENE.GC[rownames(PROCESSED_COUNTS$filteredExprMatrix), 'percentage_gc_content'],
                          lengths = GENE.LEN[rownames(PROCESSED_COUNTS$filteredExprMatrix), 'gene.length'],
                          lengthMethod = "smooth", 
                          verbose = FALSE)

# Set gene counts in specific samples that are deviating 3 sd from other samples to 0
log.mat = apply(CQN.GENE_EXPRESSION$y+CQN.GENE_EXPRESSION$offset, 1, function(x){
  mn = mean(x, na.rm = T)
  std.dev = sd(x, na.rm = T)
  return((x < (mn-3*std.dev)) | (x > (mn+3*std.dev)))
}) %>% t
PROCESSED_COUNTS$filteredExprMatrix$counts[log.mat] = 0

# Re-compute offset for gene length and gc content
CQN.GENE_EXPRESSION = cqn(PROCESSED_COUNTS$filteredExprMatrix$counts, 
                          x = GENE.GC[rownames(PROCESSED_COUNTS$filteredExprMatrix), 'percentage_gc_content'],
                          lengths = GENE.LEN[rownames(PROCESSED_COUNTS$filteredExprMatrix), 'gene.length'],
                          lengthMethod = "smooth", 
                          verbose = FALSE)
```
Coexpression of genes 
```{r coexp1, cache=FALSE, fig.height=5, fig.width=5}
cr = cor(t(CQN.GENE_EXPRESSION$y+CQN.GENE_EXPRESSION$offset))
hist(cr, main = 'Distribution of correlation between genes', xlab = 'Correlation')
```
### Outlier analysis
Clustering of voom adjusted data (with NULL design)
```{r decompse.normalise.data1.1, fig.height=6, fig.width=8, results='asis'}
# Find principal components of expression to plot
PC <- prcomp(CQN.GENE_EXPRESSION$y+CQN.GENE_EXPRESSION$offset, scale.=T, center = T)

@@

tree = hclust(as.dist(t(CQN.GENE_EXPRESSION$y + CQN.GENE_EXPRESSION$offset)))
cols = WGCNA::labels2colors(COVARIATES.tmp);

@@

### Significant Covariates
Correlation between pca of unadjusted mRNA expression and covariates are used to find significant covariates
```{r preadj.covariates, cache=TRUE}
# Find correlation between PC's of gene expression with covariates
preAdjustedSigCovars = runPCAandPlotCorrelations(CQN.GENE_EXPRESSION$y+CQN.GENE_EXPRESSION$offset, 
                                                 COVARIATES,
                                                 'NULL design(voom-normalized)', 
                                                 isKeyPlot=TRUE, 
                                                 MIN_PVE_PCT_PC = 1)
```

@@

  # Estimate voom weights for dispersion control
  VOOM.GENE_EXPRESSION = voom(PROCESSED_COUNTS$filteredExprMatrix$counts, design=DM1$design, plot=F)

  # Fit linear model using new weights and new design
  VOOM.ADJUSTED.FIT = lmFit(CQN.GENE_EXPRESSION$y+CQN.GENE_EXPRESSION$offset,
                            design = DM1$design,
                            weights = VOOM.GENE_EXPRESSION$weights)

  # Residuals after normalisation
  RESIDUAL.GENE_EXPRESSION = residuals.MArrayLM(VOOM.ADJUSTED.FIT,
                                                CQN.GENE_EXPRESSION$y+CQN.GENE_EXPRESSION$offset)

@@

# Estimate voom weights for dispersion control
VOOM.WEIGHTS = voom(PROCESSED_COUNTS$filteredExprMatrix$counts, design=DESIGN$design, plot=F)

# Fit linear model using new weights and new design
FIT = lmFit(CQN.GENE_EXPRESSION$y+CQN.GENE_EXPRESSION$offset,
            design = DESIGN$design,
            weights = VOOM.WEIGHTS$weights)

@@

# Calculate the number of surrogate variables
NUM.SV = sva::num.sv(CQN.GENE_EXPRESSION$y+CQN.GENE_EXPRESSION$offset, 
                     MODEL1, method = 'be', seed = 123456, B = 30)

# Estimate surrogate variables
SURR.VAR = sva::sva(CQN.GENE_EXPRESSION$y+CQN.GENE_EXPRESSION$offset, 
                    MODEL1, MODEL0, n.sv = NUM.SV, B = 30)$sv

@@

# Re-estimate voom weights
VOOM.SVA.WEIGHTS = voom(PROCESSED_COUNTS$filteredExprMatrix$counts, 
                        design = cbind(MODEL1, SURR.VAR), plot=F)

# Fit linear model using new weights and new design
VOOM.SVA.FIT = lmFit(CQN.GENE_EXPRESSION$y+CQN.GENE_EXPRESSION$offset, 
                     design = cbind(MODEL1, SURR.VAR),
                     weights = VOOM.SVA.WEIGHTS$weights)

# Residuals after normalisation
RESIDUAL.SVA.GENE_EXPRESSION = residuals.MArrayLM(VOOM.SVA.FIT,
                                                  CQN.GENE_EXPRESSION$y+CQN.GENE_EXPRESSION$offset)
```

@@

# Estimate voom weights for dispersion control
VOOM.NET.WEIGHTS = voom(PROCESSED_COUNTS$filteredExprMatrix$counts, 
                        design=DESIGN.NET, plot=F)

# Fit linear model using new weights and new design
VOOM.NET.FIT = lmFit(CQN.GENE_EXPRESSION$y+CQN.GENE_EXPRESSION$offset,
                     design = DESIGN.NET,
                     weights = VOOM.NET.WEIGHTS$weights)

# Residuals after normalisation
RESIDUAL.NET.GENE_EXPRESSION = residuals.MArrayLM(VOOM.NET.FIT,
                                                  CQN.GENE_EXPRESSION$y+CQN.GENE_EXPRESSION$offset)
```

@@

  normalizationType	= 'CQN'
)

# Code
CODE <- Folder(name = "ROSMAP Reprocessed RNASeq (Normalised)", parentId = parentId)
annotations(CODE) = all.annotations
CODE <- synStore(CODE)

@@

# Store logCPM
CQN.GENE_EXPRESSION$y %>%
  rownameToFirstColumn('ensembl_gene_id') %>%
  write.table(file = 'ROSMAP_DLPFC_logCPM.tsv', sep = '\t', row.names=F, quote=F)
LCOUNT_OBJ = File('ROSMAP_DLPFC_logCPM.tsv', name = 'Counts (filtered logCPM)', parentId = CODE$properties$id)
annotations(LCOUNT_OBJ) = annotations(CODE)
annotations(LCOUNT_OBJ)$dataSubType = 'filteredLCPM'
LCOUNT_OBJ = synStore(LCOUNT_OBJ, activity = synGetActivity(COV_OBJ$properties$id))

# Store cqn offsets
CQN.GENE_EXPRESSION$offset %>%
  rownameToFirstColumn('ensembl_gene_id') %>%
  write.table(file = 'ROSMAP_DLPFC_offset.tsv', sep = '\t', row.names=F, quote=F)
OFFSET_OBJ = File('ROSMAP_DLPFC_offset.tsv', name = 'Gene length and GC content offset', parentId = CODE$properties$id)
annotations(OFFSET_OBJ) = annotations(CODE)
annotations(OFFSET_OBJ)$dataSubType = 'offset'
OFFSET_OBJ = synStore(OFFSET_OBJ, activity = synGetActivity(COV_OBJ$properties$id))

>>>>

---
title: "Covariate analysis of ROSMAP reprocessed counts (with TMM normalisation and covariates adjusted)"
author: "Thanneer Perumal"
date: "`r date()`"
output: html_document
---
```{r knit2synapse, eval=FALSE}
library(synapseClient)
library(knit2synapse) # get the package from devtools::install_github('Sage-Bionetworks/knit2synapse')

@@

knit2synapse::knitToFolderEntity(file = "./ROSMAP_REPROCESSED_TMM.Rmd",
                                 parentId = "syn8403743",
                                 entityName = 'ROSMAP Reprocessed RNASeq (Normalised - With out gene length and gc content adjustment)')
```

@@

library(ggplot2)
library(reshape2)
library(limma)
library(gplots)
library(WGCNA)
library(psych)
library(edgeR)
library(biomaRt)
library(RColorBrewer)

@@

thisFileName <- 'ROSMAP_REPROCESSED_TMM.Rmd'

@@

# Sex
vcd::mosaic(~ msex + cogdx, data = COVARIATES)
```
### CPM Normalisation
Preprocess counts matrix using TMM normalisation

@@

### Normalisation (with NULL)
Initial normalisation usign cqn and dispersion control with voom (with NULL design)
```{r initial.voom.normalisation, fig.height=5, fig.width=5}
# Normalise using TMM
TMM.GENE_EXPRESSION = calcNormFactors(PROCESSED_COUNTS$filteredExprMatrix, method = 'TMM')
VOOM.GENE_EXPRESSION = voom(TMM.GENE_EXPRESSION)

# Set gene counts in specific samples that are deviating 3 sd from other samples to 0
log.mat = apply(VOOM.GENE_EXPRESSION$E, 1, function(x){
  mn = mean(x, na.rm = T)
  std.dev = sd(x, na.rm = T)
  return((x < (mn-3*std.dev)) | (x > (mn+3*std.dev)))
}) %>% t
PROCESSED_COUNTS$filteredExprMatrix$counts[log.mat] = 0

# Re-normalise using TMM
TMM.GENE_EXPRESSION = calcNormFactors(PROCESSED_COUNTS$filteredExprMatrix, method = 'TMM')
VOOM.GENE_EXPRESSION = voom(TMM.GENE_EXPRESSION)
```
Coexpression of genes 
```{r coexp1, cache=FALSE, fig.height=5, fig.width=5}
cr = cor(t(VOOM.GENE_EXPRESSION$E))
hist(cr, main = 'Distribution of correlation between genes', xlab = 'Correlation')
```
### Outlier analysis
Clustering of voom adjusted data (with NULL design)
```{r decompse.normalise.data1.1, fig.height=6, fig.width=8, results='asis'}
# Find principal components of expression to plot
PC <- prcomp(VOOM.GENE_EXPRESSION$E, scale.=T, center = T)

@@

tree = hclust(as.dist(t(VOOM.GENE_EXPRESSION$E)))
cols = WGCNA::labels2colors(COVARIATES.tmp);

@@

### Significant Covariates
Correlation between pca of unadjusted mRNA expression and covariates are used to find significant covariates
```{r preadj.covariates, cache=TRUE}
# Find correlation between PC's of gene expression with covariates
preAdjustedSigCovars = runPCAandPlotCorrelations(VOOM.GENE_EXPRESSION$E, 
                                                 COVARIATES,
                                                 'NULL design(voom-normalized)', 
                                                 isKeyPlot=TRUE, 
                                                 MIN_PVE_PCT_PC = 1)
```

@@

  # Estimate voom weights for dispersion control
  VOOM.GENE_EXPRESSION = voom(TMM.GENE_EXPRESSION, design=DM1$design, plot=F)

  # Fit linear model using new weights and new design
  VOOM.ADJUSTED.FIT = lmFit(VOOM.GENE_EXPRESSION)

  # Residuals after normalisation
  RESIDUAL.GENE_EXPRESSION = residuals.MArrayLM(VOOM.ADJUSTED.FIT,
                                                VOOM.GENE_EXPRESSION$E)

@@

# Estimate voom weights for dispersion control
VOOM.WEIGHTS = voom(TMM.GENE_EXPRESSION, design=DESIGN$design, plot=F)

# Fit linear model using new weights and new design
FIT = lmFit(VOOM.WEIGHTS)

@@

# Calculate the number of surrogate variables
NUM.SV = sva::num.sv(VOOM.WEIGHTS$E, MODEL1, method = 'be', seed = 123456, B = 30)

# Estimate surrogate variables
SURR.VAR = sva::sva(VOOM.WEIGHTS$E, MODEL1, MODEL0, n.sv = NUM.SV, B = 30)$sv

@@

# Re-estimate voom weights
VOOM.SVA.WEIGHTS = voom(TMM.GENE_EXPRESSION, 
                        design = cbind(MODEL1, SURR.VAR), plot=F)

# Fit linear model using new weights and new design
VOOM.SVA.FIT = lmFit(VOOM.SVA.WEIGHTS)

# Residuals after normalisation
RESIDUAL.SVA.GENE_EXPRESSION = residuals.MArrayLM(VOOM.SVA.FIT,
                                                  VOOM.SVA.WEIGHTS$E)
```

@@

# Estimate voom weights for dispersion control
VOOM.NET.WEIGHTS = voom(TMM.GENE_EXPRESSION, design=DESIGN.NET, plot=F)

# Fit linear model using new weights and new design
VOOM.NET.FIT = lmFit(VOOM.NET.WEIGHTS)

# Residuals after normalisation
RESIDUAL.NET.GENE_EXPRESSION = residuals.MArrayLM(VOOM.NET.FIT,
                                                  VOOM.NET.WEIGHTS$E)
```

@@

  normalizationType	= 'TMM'
)

# Code
CODE <- Folder(name = "ROSMAP Reprocessed RNASeq (Normalised - With out gene length and gc content adjustment)", parentId = parentId)
annotations(CODE) = all.annotations
CODE <- synStore(CODE)

@@

# Store logCPM
VOOM.NET.WEIGHTS$E %>%
  rownameToFirstColumn('ensembl_gene_id') %>%
  write.table(file = 'ROSMAP_DLPFC_logCPM.tsv', sep = '\t', row.names=F, quote=F)
LCOUNT_OBJ = File('ROSMAP_DLPFC_logCPM.tsv', name = 'Counts (filtered logCPM)', parentId = CODE$properties$id)
annotations(LCOUNT_OBJ) = annotations(CODE)
annotations(LCOUNT_OBJ)$dataSubType = 'filteredLCPM'
LCOUNT_OBJ = synStore(LCOUNT_OBJ, activity = synGetActivity(COV_OBJ$properties$id))
